A Fast and Tight Heuristic for A* in Road Networks

### 2021-08-20 paper/ch_potentials.tex
writing on penalty

@@

Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a fast heuristic with tight estimates.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms for a multitude of features.
In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
% The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
%
CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
For this, a faster, second preprocessing phase is introduced.
It can be run regularly to update weights.
In theory, this enables the integration of live traffic and user preferences.
In practice, live traffic feed data is imperfect.
Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{dss-tarrn-18}.
CRP also supports turn costs.
Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
On continental sized networks, TD-CRP can only compute approximate shortest distances (rather than paths).
%
In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
CCH also has a second preprocessing phase where weights can be altered.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
% Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.
%
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every extension the flexibility available at query time is fairly limited.
Combining these hierarchical extensions is an unsolved problem.

@@

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\operatorname{dist}_w(s,t)$ between $s$ and $t$.

@@

Dijkstra's algorithm can be used to solve this point to point shortest path problem.
It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $d[v] = \min(d[v], d[u] + w(uv))$.
If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
We call this a \emph{backward search}.
Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
The searches can terminate when the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$.

\subsection{A*}

A* is a goal-directed variant of Dijkstra's algorithm.
% A* uses a \emph{heuristic} to guide the search towards the goal.
It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
The priority queue is then ordered by $d[v] + h_t(v)$.
Thus, nodes closer to the target a explored earlier.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A*'s running time crucially depends on how tight this estimate is.

A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights} $w_{h_t}$.
This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
A heuristic function is called \emph{feasible} if $w_{h_t} \geq 0$ for all edges.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
		}
	}
	\Return{$P[x]$}\;
}
\caption{Lazy RPHAST algorithm}
\label{algo:pot}
\end{algorithm2e}

@@

To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i,x)$ the distance $D[x]$.
Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i,x) + D[x]\}$.
As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
% TODO consistent uv vs u,v
% A* with this heuristic is the basic CH-Potentials algorithm.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
			$\Dist(P[x])$\;
		}
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
		}
	}
	\Return{$P[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}

@@

Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
The Bucket CH selection phase works as follows:
For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
These buckets can be populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
Now we can utilize Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) to compute distances from a given source to \emph{all} targets at once.
We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
% TODO formatting
After calling the adjusted ComputeAndMemoizeDist for each source, we know the shortest distances between each pair of source and target node.
This way, we explore the search space of the sources only once.

@@

The CH-Potentials heuristic is a straightforward application of the Lazy RPHAST algorithm.
In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
Then, the heuristic function $h_t(x)$ is implemented by call the ComputeAndMemoizeDist for node $x$.

@@

Bidir A* not as straightforward which one might expect.
As pointed out in~\cite{gh-cspas-05}, one can not simply use a forward potential $h_t$ for the forward search and a backward potential $h_s$ for the reverse search.
Since these searches would operate on different reduced graphs, it is unclear when the searches can terminate.
To the best of our knowledge, no better stopping criterion exists than running both searches until the minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
That is the same stopping crit as for a unidir search..
symmetric
Thus, the symmetric bidirectional algorithm performs double the work than a simple unidir search.
can be slightly improved through pruning $\overrightarrow{D}[x] + h_t(x) > \mu$.
Still not competitive

average potential - reduced graph with average of reduced edge weights
but average pot in each dir worse than orig pot for dir
also overhead

we propose an improved pruning crit, one which takes the search progress into account (through the min queue keys), which makes the simpler symmetric approach competitive or even faster than the average pot
If $x$ was not yet settled by reverse search with minimum queue key $\overleftarrow{k}$ then
$\operatorname{dist}(x,t) \geq \overleftarrow{k} - h_s(x)$.
Thus, if $\overrightarrow{D}[x] + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.

% direction switch strats!

@@

We modify A* by processing low degree nodes consecutively without pushing them into the queue.
Our algorithm uses the undirected degree $d(x)$ of a node $x$.
Formally, $d(x)$ is the number of nodes $y$ such that $(x,y)\in E$ or $(y,x)\in E$.

@@

Our algorithm differs from A* when removing a node $x$ from the queue.
A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
Our algorithm, however, behaves differently, if $d(y)\le 2$.
Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
Instead, it iteratively tries to reduce all $D[y_i]$.
If it does not reach $z$, then only $D$ is modified but no queue action is performed.
If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.

@@

We can also skip some degree three nodes.
Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.

@@

% A* is a flexible algorithm.
% Many extended problems can therefore be solved using the CH-Potentials framework.
We start by establishing a common formal framework for the use of CH-Potentials.
We describe some extended routing problems and how to apply CH-Potentials to them.
Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.

@@

In this paper, we consider different applications, with slightly different problem models.
The goal is always to quickly answer many shortest path queries.
For the purpose of describing our framework, we establish a shared notation:
Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
However, this is not a requirement.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
However, values of $w_q$ might change between queries.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.

To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
The exact implementation of this A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
In contrast, the preprocessing phase remains the same for all applications.

Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
By requirement and because of the triangle inequality the following holds:
\[
w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
\]
Thus, A* will always determine the correct shortest distances.

@@

% Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
w_\ell(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i,v_{i+1})
\]
The objective is to find a path between two given edges with minimum turn-aware weight.
The first term $w_\ell(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
It can thus be ignored during optimization.

>>>>

\newcommand*{\dist}{\mathcal{D}}

@@

Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is function that maps a node to an estimate of the remaining distance to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a fast heuristic with tight estimates.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms for a multitude of features.
In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.

A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.

CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
For this, a faster, second preprocessing phase is introduced.
It can be run regularly to update weights.
In theory, this enables the integration of live traffic and user preferences.
In practice, live traffic feed data is imperfect.
Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{dss-tarrn-18}.
CRP also supports turn costs.
Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
On continental sized networks, TD-CRP can only compute approximate shortest distances (rather than paths).

In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
CCH also has a second preprocessing phase where weights can be altered.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
% Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.

Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.

Another feature which has received quite some attention is the computation alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13}.
Even formally defining what makes a good alternative route is non-trivial~\cite{bdgs-argrn-11}.
A variety of paradigms exists.
Competitive approaches require careful adjustment of their respective hierarchical speedup technique for the targeted paradigm.
In~\cite{krs-eepma-13}, the CRP customization is utilized to repeatedly compute shortest paths on graphs with slightly modified weights.
In~\cite{k-hdara-13}, partial shortest path trees are extracted from CH search spaces to find via-node candidates for alternative routes.
In~\cite{barth2019alternative}, a multi-criteria CH is extended to compute alternative routes with regards to multiple edge weight functions.

While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every extension the flexibility available at query time is fairly limited.
Combining these hierarchical extensions is an unsolved problem.

@@

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.

@@

Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this point to point shortest path problem.
It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $d[v] = \min(d[v], d[u] + w(uv))$.
If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
We call this a \emph{backward search}.
Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
Typically, the searches are interleaved by alternating settling a node from each direction.
Another common approach is to always settle a node from the direction with the smaller minimal queue key.
Theoretically any direction interleaving strategy is possible.
The searches can also be run in parallel in two threads.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

\subsection{A*}\label{sec:a_star}

A* is a goal-directed variant of Dijkstra's algorithm.
% A* uses a \emph{heuristic} to guide the search towards the goal.
It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
The priority queue is ordered by $d[v] + h_t(v)$.
Thus, nodes closer to the target a explored earlier.

A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
A heuristic function is called \emph{feasible} if $w_{h_t} \geq 0$ for all edges.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Lazy RPHAST algorithm}
\label{algo:pot}
\end{algorithm2e}

@@

To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i x)$ the distance $D[x]$.
Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i x) + D[x]\}$.
As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
% A* with this heuristic is the basic CH-Potentials algorithm.

@@

\begin{algorithm2e}
\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
\KwData{$P[x]$: parent node of $x$ in the elimination tree}
\SetKwFunction{Dist}{ComputeAndMemoizeDist}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Dist{$x$}}{
	\If{$D[x] = \bot$}{
		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
			$\Dist(P[x])$\;
		}
		$D[x]\leftarrow B[x]$\;
		\For{all up edges $xy$ in $G^+_\ell$}{
                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
		}
	}
	\Return{$D[x]$}\;
}
\caption{Elimination tree based Lazy RPHAST algorithm}
\label{algo:cch_pot}
\end{algorithm2e}

@@

Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
The Bucket CH selection phase works as follows:
For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
These buckets can be populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
Now we can utilize Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) to compute distances from a given source to \emph{all} targets at once.
We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
This way, we explore the search space of the sources only once.

@@

The CH-Potentials heuristic is a straightforward application of the Lazy RPHAST algorithm.
In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
The heuristic function $h_t(x)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for node $x$.

@@

Since bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm, a bidirectional variant of A* also seems desirable.
However, as pointed out in~\cite{gh-cspas-05}, this is not as straightforward as one might expect.
One would like to use a forward potential $\overrightarrow{h}_t$ in the forward search and a backward potential $\overleftarrow{h}_s$ (computing distances in the reversed graph) in the reverse search.
The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
Thus, in the equivalent bidirectional Dijkstra search each direction would use different graph.
This would break the bidirectional Dijkstra stopping criterion.
To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
But this is the same stopping criterion that a unidirectional A* search has.
Thus, this straightforward bidirectional A* performs about twice the work than a unidirectional A*.
This can be slightly improved by introducing pruning.
Nodes, for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e. their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
This approach is denoted as the \emph{symmetric} bidirectional A*.

To overcome this problem the authors of propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
The reduced graph of the average potential is the graph with the average weights of the reduced graphs of the forward and backward potentials.
The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
This potential allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
Additionally, the pruning from the symmetric approach can be utilized with the original potentials.
To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
Still, it has two downsides.
First, evaluating the average potential requires evaluating the original heuristics for both directions.
This can cause some overhead depending on the heuristics.
Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.

We propose an improved pruning criterion for the symmetric approach.
It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
As our experiments show, this makes the simpler symmetric approach at least competitive and sometimes even faster than the average potential.
Recall that the queue is ordered by $D[x] + h(x)$.
We use the minimal queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$.
If a node $x$ was not yet settled by backward search then $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
The pruning for the backwards search works analogously.

@@

We modify A* by processing low degree nodes consecutively without pushing them into the queue.
Our algorithm uses the undirected degree $d(x)$ of a node $x$.
Formally, $d(x)$ is the number of nodes $y$ such that $(xy)\in E$ or $(yx)\in E$.

@@

Our algorithm differs from A* when removing a node $x$ from the queue.
A* iterates over the outgoing arcs $(xy)$ of $x$ and tries to reduce $D[y]$ by relaxing $(xy)$.
If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
Our algorithm, however, behaves differently, if $d(y)\le 2$.
Our algorithm determines the longest degree two chain of nodes $x y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
Instead, it iteratively tries to reduce all $D[y_i]$.
If it does not reach $z$, then only $D$ is modified but no queue action is performed.
If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.

@@

We can also skip some degree three nodes.
Denote by $x y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z a_1,\ldots,a_p b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.

@@

% A* is a flexible algorithm.
% Many extended problems can therefore be solved using the CH-Potentials framework.
We start by establishing a common formal framework for the use of CH-Potentials.
Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.

@@

We consider a variety different applications, with slightly different problem models.
The goal is always to quickly answer many shortest path queries.
For the purpose of describing our framework, we establish a shared notation:
Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
In the simplest case, $w_q$ will be scalar edge weights.
However, this is not a requirement.
It can be any function that computes a weight for an edge.
This function can also take additional parameters from the state of the search.
For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
However, values of $w_q$ might change between queries.
In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.

To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.

The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
The exact implementation of this A* search depends on the application.
Our approach only provides the heuristic $h_t$ for the A* search.
In contrast, the preprocessing phase remains the same for all applications.

Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
By requirement and because of the triangle inequality the following holds:
\[
w_q(u v) - h_t(u) + h_t(v) \geq \dist_\ell(\phi(u), \phi(v)) - \dist_\ell(\phi(u), \phi(t)) + \dist_\ell(\phi(v), \phi(t)) \geq 0
\]
Thus, A* will always determine the correct shortest distances.

@@

% Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i v_{i+1})
\]
The objective is to find a path between two given edges with minimum turn-aware weight.
The first term $w_\ell(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
It can thus be ignored during optimization.

We solve this problem by constructing a \emph{turn-expanded} graph as $G_q$.
Edges in the input graph $G_\ell$ correspond to \emph{expanded nodes} in $G_q$.
For every pair of incident edges $x y$ and $y z$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_\ell(y z)$.
A sequence of expanded nodes in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
The weight of a path in $G_q$ is equal to the turn-aware weight of the corresponding path in $G_\ell$ minus the irrelevant $w_\ell(v_1,v_2)$ term.
Thus, the turn-aware routing problem can be solved by searching for shortest paths in $G_q$.

In this scenario, preprocessing and query use different graphs $G_\ell$ and $G_q$.
We define the node mapping function $\phi$ as $\phi(x y) = y$.
Obviously, $w_q(xy, yz) = w_t(x,y,z) + w_\ell(y z) \geq \dist_\ell(\phi(x y), \phi(y z))$ and this approach yields a feasible heuristic.
Sadly, the undirected graph underlying $G_q$ is always biconnected, if the input graph is strongly connected.
The optimization described in Section~\ref{sec:largested-biconnected-component} is therefore ineffective.
With this setup, CH-Potentials support turns without requiring turn information in the CH.
